---
title: 'Stat Spotlight: Eviction Filings Are Soaring in Oklahoma County'
author: 'Bean D. Cat'
date: '2022-06-16'
slug: []
categories: []
tags: []
draft: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(ojodb)
library(plotly)
library(tidyverse)

knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      echo = FALSE,
                      fig.align = 'center')

ds <- ojo_civ_cases(case_types = "SC", file_years = c(2019:2022)) |>
  filter(str_detect(description, "FORCI")) |>
  collect() 

```


```{r graph, fig.cap="Eviction filings are on the rise."}

monthly_filings <- ds |>
  count(district, month_filed = floor_date(date_filed, "months")) |>
  filter(month_filed < ymd("2022-06-01"))
  
graph <- monthly_filings |>
  ggplot(aes(x = month_filed,
             y = n,
             color = district)) +
  geom_line() +
  scale_y_continuous(limits = c(0, NA)) +
  guides(color = "none") +
  ggthemes::theme_fivethirtyeight()

ggplotly(graph)

```

---

Eviction filings across the state have been rising steadily since the onset of the COVID-19 Pandemic, despite eviction moratoriums and assistance programs aimed at preventing people from losing their homes due to the crisis. In Oklahoma County in particular, though, it seems that the dam may have finally broken -- in February of 2022, Oklahoma County landlords filed `r monthly_filings |> filter(district == "OKLAHOMA", month_filed == ymd("2022-02-01")) |> pull(var = n) |> format(big.mark = ",")` eviction cases, which is more than any month on record.




